The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2020-06-15 Fsing/FrobeniusThresholdsPaper/response_to_report_C_on_v2.txt

Consider renaming "fpt" to "pureThreshold" and "isFPT" to "isPureThreshold".
(The "F" for "Frobenius" can be assumed.)

*****************************************************************************

Consider removing redundant parentheses...

-----------------------------------------------------------------------------
We believe we have now removed the vast majority of redundant parentheses.
-----------------------------------------------------------------------------

@@

you can write

		if not isWellDefined h then

Fix all such comparisons with "true" and "false".

-----------------------------------------------------------------------------
Done.
-----------------------------------------------------------------------------

@@

*****************************************************************************

The value of an "if" expression is "null" if the predicate is false, so this
can be omitted:

    else (
        null
    )

-----------------------------------------------------------------------------
We'd rather keep the "else null", to make clear the intention to return null.
-----------------------------------------------------------------------------

@@

The newline characters in verbose messages such as this one

        if o.Verbose then print "\nVerifying if special algorithms apply...";

resulting in double-spaced output, wasting space on the screen and on the pages
of the article.

-----------------------------------------------------------------------------
Those were removed.
-----------------------------------------------------------------------------

@@

In

	Consider a Veronese subring (which is étale in codimension 1)

the use of "étale" is incorrect -- it should be applied to ring homomorphism,
not to a ring.

-----------------------------------------------------------------------------
All of the above were fixed.
-----------------------------------------------------------------------------

@@

This

	 the condition that the Frobenius morphism splits as a map of rings

is ambiguous.  Do you mean "is a split surjection" or "is a split injection"?

-----------------------------------------------------------------------------
We removed the ambiguity, saying it is a split injection. 
-----------------------------------------------------------------------------

@@

*****************************************************************************

It would be better if comments in the displayed example code, such as "uses
binary search (default)", were in the surrounding text, because then complete
sentences could be used, and they would be read by the reader in the normal
sequence.  For this purpose, displays could be split, to allow comments in the
text between the two parts.

-----------------------------------------------------------------------------
Those comments were removed and incorporated in surrounding text. 
-----------------------------------------------------------------------------

@@

*****************************************************************************

The displayed example output seems to be a diffferent color, which is
disconcerting.

-----------------------------------------------------------------------------
The displayed examples are now shown in black.
-----------------------------------------------------------------------------

@@

*****************************************************************************

The displayed example output would look better if it were indented.

-----------------------------------------------------------------------------
They are now indented.
-----------------------------------------------------------------------------

### 2020-06-15 Fsing/FrobeniusThresholdsPaper/FrobeniusThresholds.tex
Made changes to the paper suggested by referee

\usepackage{fancyvrb}
\DefineVerbatimEnvironment%
{MyVerbatim}{Verbatim}
{formatcom=\color{Violet}}

@@

This paper describes the \emph{Macaulay2} package \emph{FrobeniusThresholds}\footnote{This package was previously named \emph{FThresholds}} \cite{M2, FThresholdsPackage}, which provides tools for computing or estimating certain fundamental invariants in positive characteristic commutative algebra, namely \emph{$F$-pure thresholds}, \emph{$F$-thresholds}, and \emph{$F$-jumping exponents}.
Recall that a ring of prime characteristic $p>0$ is \emph{$F$-pure} if the Frobenius map|that is, the ring endomorphism sending an element to its $p$-th power|is a pure morphism;  under natural geometric hypotheses, this is equivalent to the condition that the Frobenius morphism splits as a map of rings.
The concept of $F$-purity has proven to be important in commutative algebra, and has a rich history.
Indeed, it first appeared in the work of Hochster and Roberts to study local cohomology \cite{HochsterRobertsFrobeniusLocalCohomology}, was compared with rational singularities in  \cite{FedderFPureRat}, and was used to study global properties of Schubert varieties in \cite{MehtaRamanathanFrobeniusSplittingAndCohomologyVanishing}.

@@

The user can specify how the search is approached through the option \texttt{Search}, which can take two values: 
 \texttt{Binary} (the default value) and \texttt{Linear}.

@@

i9 : time nu(2, m, m^2) -- uses binary search (default)
     -- used 1.82479 seconds

o9 = 97

i10 : time nu(2, m, m^2, Search => Linear) -- but linear search gets luckier
     -- used 0.597035 seconds

o10 = 97
\end{MyVerbatim}
}
%\medskip
\smallskip

@@

% \medskip
\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i12 : R = ZZ/11[x,y,z];

@@

i14 : time nu(3, f) -- ContainmentTest is set to FrobeniusRoot, by default
     -- used 0.153691 seconds

o14 = 1209

i15 : time nu(3, f, ContainmentTest => StandardPower)
     -- used 10.1343 seconds

o15 = 1209
\end{MyVerbatim}
}
%\medskip

@@

As pointed out in the introduction, if $f \in \mathfrak{m}$, the values $\nu^{\mathfrak{m}}_f(p^e)$ can be recovered from the $F$-pure threshold of $f$.
This is used to speed up computations for certain polynomials whose $F$-pure thresholds can be computed quickly via specialized algorithms or formulas, namely diagonal polynomials, binomials, forms in two variables, and products of factors in simple normal crossing (see Section~\ref{sec.FPT}).
This feature can be disabled by setting the option \texttt{UseSpecialAlgorithms} (default value \texttt{true}) to \texttt{false}.\footnote{In Section~\ref{ss: UseSpecialAlgorithms} we discuss a couple of situations in which this may be desirable.}

@@

i21 : f = x^3 + y^4 + z^5; -- a diagonal polynomial

@@

The \texttt{fpt} function has an option called \texttt{UseSpecialAlgorithms}, which, when set to \texttt{true} (its default value), tells \texttt{fpt} to first check whether $f$ is a diagonal polynomial, a binomial, a form in two variables, or a product of factors in simple normal crossing, in that order.
When $f$ is a diagonal polynomial, a binomial, or a form in two variables, algorithms of Hern\'andez \cite{HernandezFInvariantsOfDiagonalHyp, HernandezFPureThresholdOfBinomial} or Hern\'andez and Teixeira \cite{HernandezTeixeiraFThresholdFunctions} are executed to compute the $F$-pure threshold.
%Algorithms for computing $F$-pure thresholds of binomials also appear in \cite{ShibutaTakagiLCThresholds}, though they are only valid in certain special characteristics.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i4 : fpt(x^17 + y^20 + z^24) -- a diagonal polynomial

@@

o4 : QQ

i5 : fpt(x^2*y^6*z^10 + x^10*y^5*z^3) -- a binomial

@@

o5 : QQ

i6 : R = ZZ/5[x,y];

i7 : fpt(x^2*y^6*(x + y)^9*(x + 3*y)^10) -- a form in two variables

@@

\noindent The algorithm for computing the $F$-pure threshold of a binary form $f$ requires factoring $f$ into linear forms, and that may be difficult or impossible when that factorization occurs in a Galois field of excessively large order.
This is a situation when the user will want to set the option \texttt{UseSpecialAlgorithms} to \texttt{false}.
However, when a factorization is already known, instead of passing $f$ to \texttt{fpt}, the user can pass a list containing all the pairwise coprime linear factors of $f$ to \texttt{fpt}, and a list containing their respective multiplicities.

@@

If \texttt{UseSpecialAlgorithms} is set to \texttt{true} and $f$ does not fall into any of the aforementioned cases, then the function \texttt{fpt} next calls \texttt{isSimpleNormalCrossing(f)} (see Section~\ref{subsec.SNC}) to check whether the polynomial $f$ is (locally, at the origin) a product of factors that are in simple normal crossing, in which case the $F$-pure threshold is simply the reciprocal of the largest multiplicity occurring in that factorization.  
Note that the function \texttt{factor} is called whenever \texttt{isSimpleNormalCrossing} is used, and that can sometimes make the verification slow.  The user can avoid this by setting \texttt{UseSpecialAlgorithms} to \texttt{false}.

@@

With these estimates in hand, the subroutine \texttt{guessFPT} is called to make some ``educated guesses" in an attempt to identify the $F$-pure threshold within this interval, or at least narrow down this interval to produce improved estimates.  The number of ``guesses" is controlled by the option \texttt{Attempts}, which conservatively defaults to \texttt{3}.  If \texttt{Attempts} is set to \texttt{0}, then \texttt{guessFPT} is bypassed. If  \texttt{Attempts} is set to at least \texttt{1}, then a first check is run to verify whether the right-hand endpoint $(\nu+1)/p^e$ of the above interval \eqref{estimating-interval: e} is the $F$-pure threshold.  We illustrate this below.

@@

i10 : fpt(f, Attempts => 0) -- a bad estimate

@@

o10 : List

i11 : fpt(f, Attempts => 0, DepthOfSearch => 3) -- a better estimate

@@

o11 : List

i12 : fpt(f, Attempts => 1, DepthOfSearch => 3) -- the right-hand endpoint
      (ν+1)/p^e is the F-pure threshold

@@

o14 : List

i15 : fpt(f, Attempts => 2, DepthOfSearch => 3) -- the left-hand endpoint
      ν/(p^e-1) is the F-pure threshold

@@

o24 : List

i25 : numeric fpt(f, DepthOfSearch => 3, FinalAttempt => true)
      -- FinalAttempt improves the estimate slightly

@@

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i29 : fpt(f, DepthOfSearch => 3, FinalAttempt => true, Verbose => true)

Starting fpt ...

fpt is not 1 ...

Verifying if special algorithms apply...

Special fpt algorithms were not used ...

ν has been computed: ν = nu(3,f) = 17 ...

fpt lies in the interval [ν/(p^e-1),(ν+1)/p^e] = [17/124,18/125] ...

Starting guessFPT ...

The right-hand endpoint is not the fpt ...

The left-hand endpoint is not the fpt ...

guessFPT narrowed the interval down to (7/50,18/125) ...

Beginning F-signature computation ...

First F-signature computed: s(f,(ν-1)/p^e) = 793/15625 ...

Second F-signature computed: s(f,ν/p^e) = 342/15625 ...

Computed F-signature secant line intercept: 8009/56375 ...

F-signature intercept is an improved lower bound;
Using F-regularity to check if it is the fpt ...

The new lower bound is not the fpt ...

fpt failed to find the exact answer; try increasing the value of
    DepthOfSearch or Attempts.

fpt lies in the interval (8009/56375,18/125).

@@

\subsection{The function \texttt{isSimpleNormalCrossings}} \label{subsec.SNC}
As mentioned earlier, the function \texttt{isSimpleNormalCrossings} verifies whether a polynomial $f$ is a product of factors in simple normal crossing.
Suppose that $f$ has factorization $f_1^{a_i} f_2^{a_2} \cdots f_n^{a_n}$.  Recall that its factors $f_i$ are said to be in
\emph{simple normal crossing} at a point if, locally, they form part of a regular system of parameters.  The function \texttt{isSimpleNormalCrossings} determines whether $f$ has simple normal crossings at the origin by computing the Jacobian matrix of each subset of $\{ f_1, \ldots, f_n \}$ (evaluated at the origin), and checking that these matrices have the expected rank, and that these subsets generate ideals of the appropriate height.

@@

The function \texttt{isSimpleNormalCrossing} is exposed to the user, so can be used independent of any $F$-pure threshold calculation.
If the user sets its option \texttt{AtOrigin} to \texttt{false} (its default value is \texttt{true}), then the function checks whether the $f_i$ are in simple normal crossing \emph{everywhere}, which can be much slower, since Jacobian ideals are computed.

@@

i38 : f = (y - (x - 1)^2)*y^2; --SNC at the origin, but not globally

>>>>

\usepackage{fancyvrb}
\DefineVerbatimEnvironment%
{MyVerbatim}{Verbatim}
{xleftmargin=5mm}
%{formatcom=\color{Violet}}

@@

This paper describes the \emph{Macaulay2} package \emph{FrobeniusThresholds}\footnote{This package was previously named \emph{FThresholds}} \cite{M2, FThresholdsPackage}, which provides tools for computing or estimating certain fundamental invariants in positive characteristic commutative algebra, namely \emph{$F$-pure thresholds}, \emph{$F$-thresholds}, and \emph{$F$-jumping exponents}.
Recall that a ring of prime characteristic $p>0$ is \emph{$F$-pure} if the Frobenius map|that is, the ring endomorphism sending an element to its $p$-th power|is a pure morphism;  under natural geometric hypotheses, this is equivalent to the condition that the Frobenius morphism is a split injection of rings.
The concept of $F$-purity has proven to be important in commutative algebra, and has a rich history.
Indeed, it first appeared in the work of Hochster and Roberts to study local cohomology \cite{HochsterRobertsFrobeniusLocalCohomology}, was compared with rational singularities in  \cite{FedderFPureRat}, and was used to study global properties of Schubert varieties in \cite{MehtaRamanathanFrobeniusSplittingAndCohomologyVanishing}.

@@

The user can specify how the search is approached through the option \texttt{Search}, which can take two values: 
\texttt{Binary} (the default value) and \texttt{Linear}.
In the example below, the default search method, \texttt{Binary}, is used.

@@

i9 : time nu(2, m, m^2)
     -- used 1.82479 seconds

o9 = 97
\end{MyVerbatim}
}
%\medskip
\smallskip

\noindent
However, a linear search is faster in this case.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i10 : time nu(2, m, m^2, Search => Linear)
     -- used 0.597035 seconds

o10 = 97
\end{MyVerbatim}
}
\smallskip

@@

In the example below, \texttt{ContainmentTest} is set to its default value for polynomials, namely \texttt{FrobeniusRoot}.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i12 : R = ZZ/11[x,y,z];

@@

i14 : time nu(3, f) 
     -- used 0.153691 seconds

o14 = 1209
\end{MyVerbatim}
}
\smallskip

\noindent
If \texttt{ContainmentTest} is set to \texttt{StandardPower}, instead, the computation is significantly slower.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i15 : time nu(3, f, ContainmentTest => StandardPower)
     -- used 10.1343 seconds

o15 = 1209
\end{MyVerbatim}
}
\smallskip

@@

As pointed out in the introduction, if $f \in \mathfrak{m}$, the values $\nu^{\mathfrak{m}}_f(p^e)$ can be recovered from the $F$-pure threshold of $f$.
This is used to speed up computations for certain polynomials whose $F$-pure thresholds can be computed quickly via specialized algorithms or formulas, namely diagonal polynomials, binomials, forms in two variables, and polynomials that define simple normal crossing divisors (see Section~\ref{sec.FPT}).
This feature can be disabled by setting the option \texttt{UseSpecialAlgorithms} (default value \texttt{true}) to \texttt{false}.\footnote{In Section~\ref{ss: UseSpecialAlgorithms} we discuss a couple of situations in which this may be desirable.}

The following example shows, for a diagonal polynomial, how much faster the computation can be when special algorithms are enabled.

@@

i21 : f = x^3 + y^4 + z^5;

@@

The \texttt{fpt} function has an option called \texttt{UseSpecialAlgorithms}, which, when set to \texttt{true} (its default value), tells \texttt{fpt} to first check whether $f$ is a diagonal polynomial, a binomial, a form in two variables, or a polynomial that defines a simple normal crossing divisor, in that order.
When $f$ is a diagonal polynomial, a binomial, or a form in two variables, algorithms of Hern\'andez \cite{HernandezFInvariantsOfDiagonalHyp, HernandezFPureThresholdOfBinomial} or Hern\'andez and Teixeira \cite{HernandezTeixeiraFThresholdFunctions} are executed to compute the $F$-pure threshold.

In the following example, we compute the $F$-pure threshold of a diagonal polynomial.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i4 : fpt(x^17 + y^20 + z^24)

@@

o4 : QQ
\end{MyVerbatim}
}
\smallskip

\noindent
Next, we compute the $F$-pure threshold of a binomial.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i5 : fpt(x^2*y^6*z^10 + x^10*y^5*z^3)

@@

o5 : QQ
\end{MyVerbatim}
}
\smallskip

\noindent
Finally, we compute the $F$-pure threshold of a form in two variables.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i6 : R = ZZ/5[x,y];

i7 : fpt(x^2*y^6*(x + y)^9*(x + 3*y)^10)

@@

The algorithm for computing the $F$-pure threshold of a binary form $f$ requires factoring $f$ into linear forms, and that may be difficult or impossible when that factorization occurs in a Galois field of excessively large order.
This is a situation when the user will want to set the option \texttt{UseSpecialAlgorithms} to \texttt{false}.
However, when a factorization is already known, instead of passing $f$ to \texttt{fpt}, the user can pass a list containing all the pairwise coprime linear factors of $f$ to \texttt{fpt}, and a list containing their respective multiplicities.

@@

If \texttt{UseSpecialAlgorithms} is set to \texttt{true} and $f$ does not fall into any of the aforementioned cases, then the function \texttt{fpt} next calls \texttt{isSimpleNormalCrossing(f)} (see Section~\ref{subsec.SNC}) to check whether the polynomial $f$ defines (locally, at the origin) a simple normal crossing divisor, in which case the $F$-pure threshold is simply the reciprocal of the largest multiplicity occurring in that factorization.  
Note that the function \texttt{factor} is called whenever \texttt{isSimpleNormalCrossing} is used, and that can sometimes make the verification slow.  The user can avoid this by setting \texttt{UseSpecialAlgorithms} to \texttt{false}.

@@

With these estimates in hand, the subroutine \texttt{guessFPT} is called to make some ``educated guesses" in an attempt to identify the $F$-pure threshold within this interval, or at least narrow down this interval to produce improved estimates.  The number of ``guesses" is controlled by the option \texttt{Attempts}, which conservatively defaults to \texttt{3}.  If \texttt{Attempts} is set to \texttt{0}, then \texttt{guessFPT} is bypassed. If  \texttt{Attempts} is set to at least \texttt{1}, then a first check is run to verify whether the right-hand endpoint $(\nu+1)/p^e$ of the above interval \eqref{estimating-interval: e} is the $F$-pure threshold.

To illustrate these options, first we obtain a rather crude estimate for the $F$-threshold of a polynomial.

@@

i10 : fpt(f, Attempts => 0)

@@

o10 : List
\end{MyVerbatim}
}
\smallskip

\noindent
Increasing the depth of search, we obtain a better estimate.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i11 : fpt(f, Attempts => 0, DepthOfSearch => 3)

@@

o11 : List
\end{MyVerbatim}
}
\smallskip

\noindent
Finally, increasing the number of attempts we find that the right-hand endpoint of the above interval is the desired $F$-pure threshold.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i12 : fpt(f, Attempts => 1, DepthOfSearch => 3)

@@

o14 : List
\end{MyVerbatim}
}
\smallskip

\noindent
With \texttt{Attempts} set to \texttt{2}, we find that the left-hand endpoint of the above interval is the desired $F$-pure threshold.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i15 : fpt(f, Attempts => 2, DepthOfSearch => 3)

@@

o24 : List
\end{MyVerbatim}
}
\smallskip

\noindent
With \texttt{FinalAttempt} set to \texttt{true}, we can slightly improve this estimate.

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i25 : numeric fpt(f, DepthOfSearch => 3, FinalAttempt => true)

@@

\smallskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i29 : fpt(f, DepthOfSearch => 3, FinalAttempt => true, Verbose => true)
Starting fpt ...
fpt is not 1 ...
Verifying whether special algorithms apply...
Special fpt algorithms were not used ...
ν has been computed: ν = nu(3,f) = 17 ...
fpt lies in the interval [ν/(p^e-1),(ν+1)/p^e] = [17/124,18/125] ...
Starting guessFPT ...
The right-hand endpoint is not the fpt ...
The left-hand endpoint is not the fpt ...
guessFPT narrowed the interval down to (7/50,18/125) ...
Beginning F-signature computation ...
First F-signature computed: s(f,(ν-1)/p^e) = 793/15625 ...
Second F-signature computed: s(f,ν/p^e) = 342/15625 ...
Computed F-signature secant line intercept: 8009/56375 ...
F-signature intercept is an improved lower bound;
Using F-regularity to check if it is the fpt ...
The new lower bound is not the fpt ...
fpt failed to find the exact answer; try increasing the value of
    DepthOfSearch or Attempts.
fpt lies in the interval (8009/56375,18/125).

@@

\subsection{The function \texttt{isSimpleNormalCrossing}} \label{subsec.SNC}
As mentioned earlier, the function \texttt{isSimpleNormalCrossing} verifies whether a polynomial $f$ defines a simple normal crossing divisor.
Suppose that $f$ has factorization $f_1^{a_i} f_2^{a_2} \cdots f_n^{a_n}$.
Recall that $f$ defines a simple normal crossing divisor at a point if, locally, its factors form part of a regular system of parameters.
The function \texttt{isSimpleNormalCrossing} determines whether $f$ defines a simple normal crossing divisor at the origin by computing the Jacobian matrix of each subset of $\{ f_1, \ldots, f_n \}$ (evaluated at the origin), and checking that these matrices have the expected rank, and that these subsets generate ideals of the appropriate height.

@@

The function \texttt{isSimpleNormalCrossing} is exposed to the user, so can be used independently of any $F$-pure threshold calculation.
If the user sets its option \texttt{AtOrigin} to \texttt{false} (its default value is \texttt{true}), then the function checks whether $f$ defines a simple normal crossing divisor \emph{everywhere}, which can be much slower, since Jacobian ideals are computed.

@@

i38 : f = (y - (x - 1)^2)*y^2;
